---
title: "Advanced Supersonic Routing"
author: "David Marsh"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette  
vignette: >
  %\VignetteIndexEntry{Advanced Supersonic Routing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
## Introduction

[Supersonic Routes](SupersonicRouting.html) provides a quick end-to-end introduction to using the `himach` package and is the place to start. This vignette provides advice on more advanced use, explaining details that the introduction skates over.

Much of this vignette is optional or for occasional use, but the advice on saving and reading the cache is likely to be _essential_ for a speedy workflow. 


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
#the libraries needed for the vignette are
library(himach)
library(dplyr, quietly = TRUE, warn.conflicts = FALSE)
library(ggplot2)
library(sp)
library(sf)

```


## Cache {#cache}

`himach` uses caching to speed things up. Legs are cached in `route_cache` and arrival-departure links to airports are cached in `star_cache` (STAR is short for standard arrival route, its counterpart being SID the standard instrument departure).

You will want to save and load the cache (meaning the combination of `route_cache` and `star_cache`) as part of your standard workflow. Quite where you save it is up to you, but a set of routes is sensitive to (a) the route grid on which it is calculated (b) the list of aircraft used, and their performance. The saving function `hm_save_cache` forces you to refer to these two datasets, and uses metadata from them in the file name for the cache. 

If you change either of these, then you can use `hm_clean_cache()` to empty the cache. You will also note that if you run `find_routes` and the map has changed, or `findToCToD` and the map or aircraft have changed, then the cache will be cleared automatically.

For the vignette, we save to a temporary directory. You really don't want to do this in practice ;-) 

```{r cache saving, eval=FALSE} 
hm_clean_cache() #start without cache

# need to load some of the built-in data for this example
aircraft <- make_aircraft(warn = FALSE)
NZ_buffer_Pac <- sf::st_transform(NZ_buffer30, crs=crs_Pacific)
airports <- make_airports(crs = crs_Pacific)

options("quiet"= 2) # for a little reporting
# how long does it take with an empty cache?
system.time(
  routes <- find_route(aircraft[1, ],
                       make_AP2("NZAA", "NZDN", airports),
                       fat_map = NZ_buffer_Pac,
                       route_grid = NZ_grid,
                       ap_loc = airports)
)

# test saving of cache to a disposable directory
tmp_dir <- tempdir()
# for convenience, hm_save_cache gives the full name, including path
full_filename <- hm_save_cache("test", NZ_grid, aircraft, path = tmp_dir)

#empty cache - just to demonstrate the re-loading
# this isn't part of your normal workflow!
hm_clean_cache() 
# but normally a session will begin with loading a cache like this
hm_load_cache(full_filename)

# how long does it take with a cache?
system.time(
  routes <- find_route(aircraft[1, ],
                       make_AP2("NZAA", "NZDN", airports),
                       fat_map = NZ_buffer_Pac,
                       route_grid = NZ_grid,
                       ap_loc = airports)
)

# if you want to see a map
# map_routes(NZ_coast, routes, crs_Pacific, fat_map = NZ_buffer_Pac)
```
  
The cache just works invisibly in the background - you will notice it speeds up finding of routes no end: in that example, from 1.5s (user) to 0.04s (user) on my machine. In particular, it helps with refuelling, because the `route_cache` quickly remembers the routes from major hub airports to the main refuelling points, so they don't need to be calculated again.

Incidentally, if you add a new refuelling point, then the cache remains _valid_ because only **legs are cached, not routes**. With a new refuelling point, `find_route` will check both old legs and new (to the new refuelling points), gaining where the legs are cached, before selecting the best combination of legs to make the route.

# No-fly zones

It is not unusual for parts of the airspace to be closed, or be considered unsafe for flying. `himach` allows regions to be marked as 'avoid'. They will not feature in the grid, so routes will avoid them, with one exception: an arrival or departure airport can be inside a no-fly zone, as long as the connection point to the grid is outside. So they might more precisely be called 'no-overfly' zones.

A no-fly zone is prepared in the same way as a map of land. If specific countries are to be avoided, this is where having a country name in the geographic data comes in handy.

One essential item is the `avoid` attribute of the no-fly zone. This is used to distinguish sets of legs with different, or no, no-fly zone.

In this example, no offence is intended to the citizens of Buller District of New Zealand; it is a convenient example for showing how routes are forced to change when airspace is unavailable.

```{r createBuffer}
# using your own shp file 
# NZ_Buller <- sf::read_sf("...../territorial-authority-2020-clipped-generalised.shp") %>% 
#     filter(TA2020_V_1 == "Buller District")
# NZ_Buller_u <- sf::st_union(sf::st_simplify(NZ_Buller, dTolerance = 1000))
# NZ_Buller_buffer50 <- sf::st_union(sf::st_buffer(NZ_Buller_u, 50 * 1000))
# attr(NZ_Buller_buffer50, "avoid") <- "Buller+50km"
# the quicker version, using a built-in no fly zone
Buller_nofly <- sf::st_transform(NZ_Buller_buffer40, crs=crs_Pacific)
attr(Buller_nofly, "avoid") <- "Buller+40km" #required for correct caching

# this uses data as in the previous code chunk
aircraft <- make_aircraft(warn = FALSE)
NZ_buffer_Pac <- sf::st_transform(NZ_buffer30, crs=crs_Pacific)
airports <- make_airports(crs = crs_Pacific)


# run the same route, but with the avoid region
options("quiet"= 0) #just the progress bar
ac <- aircraft[c(1, 4), ]$id
routes <- find_routes(ac, 
                      data.frame(ADEP = "NZAA", ADES = "NZDN"),
                      aircraft, airports,
                      fat_map = NZ_buffer_Pac, 
                      route_grid = NZ_grid,
                      cf_subsonic = aircraft[3,],
                      avoid = Buller_nofly)

#this shows versions of the legs with and without no-fly
# ls(route_cache, pattern = "NZCH", envir = .hm_cache)

# create route summary
rtes <- summarise_routes(routes, airports)

# draw a basic map
map_routes(NZ_coast, routes, crs_Pacific, fat_map = NZ_buffer_Pac,
          avoid_map = Buller_nofly)

map_routes(NZ_coast, routes, show_route = "aircraft",
           crs = crs_Pacific, fat_map = NZ_buffer_Pac,
          avoid_map = Buller_nofly)

```


# Too many ocean routes: Buffer widths and projections {#bufferaccuracy}

Above 60 or 70 (North or South), the approximations used by the `st_buffer` function begin to show signs of exceeding their limits. In particular, if you're adding a 50km coastal buffer, for example, there are separations between Canadian islands which are just under 100km. Borden and Ellef Ringnes are examples. A buffer generated by `st_buffer` shows the strait between them as open water, where it should be closed. 

This can lead to over-optimistic routings: supersonic where they should not be. 

The solution is to use the links from `sf` to the `s2` package which come in more recent versions of the `sf` package. This does require you to use quite a high value for the `max_cells` parameter of `s2::s2_buffer_cells`.

```{r buffers in s2, fig.width=7, eval = FALSE}
gr <- s2::s2_data_countries(c("Greenland", "Canada", "Iceland"))
gr_buffer_s2 <- s2::s2_buffer_cells(gr, distance = 50000, max_cells = 20000) %>%
   st_as_sfc()
m_s2 <- ggplot(st_transform(gr_buffer_s2, crs_Atlantic)) + geom_sf(fill = "grey40") +
   geom_sf(data = st_transform(st_as_sfc(gr), crs_Atlantic))

sf_use_s2(FALSE) # to be sure
gr_transf <- gr %>%
   st_as_sfc() %>%
   st_transform(crs_Atlantic)
gr_t_buffer <- gr_transf %>%
   st_buffer(dist = 50000)
m_old <- ggplot(gr_t_buffer) + geom_sf(fill = "grey40") + geom_sf(data = gr_transf)

cowplot::plot_grid(m_old, m_s2, labels = c("bad", "good"),
                   ncol = 1)

```

![Incorrect and Correct Buffer](buffer_example.pdf) 

In fact, the problem of finding too many apparently over-ocean routes is broader than this. The other main contributor to this is missing islands from the map. See the comments [in the first vignette](SupersonicRouting.html#maps).

An example of this is in the same place. Some maps omit small islands (well, larger ones like Killniq down to tiny ones like Goodwin Island) at the mouth of the Hudson Strait. This affects the apparent width of the opening. Given the islands, and a 50km buffer, the Strait is _not_ open as the next example shows.

(This uses a non-CRAN, but public package of hi-resolution maps. If you don't want to load it, just note the results in the figure.)

```{r Island Example, fig.width=7, eval = FALSE}
sf::sf_use_s2(TRUE)
hires <- sf::st_as_sf(rnaturalearthhires::countries10) %>% 
  filter(NAME %in% c("Greenland", "Canada", "Iceland"))
hires_buffer_s2 <- s2::s2_buffer_cells(hires, distance = 50000, max_cells = 20000) %>%
   st_as_sfc()
m_hires <- ggplot(st_transform(hires_buffer_s2, crs_Atlantic)) + 
  geom_sf(fill = "grey40") +
   geom_sf(data = st_transform(hires, crs_Atlantic))

cowplot::plot_grid(m_s2, m_hires, labels = c("good", "better"),
                   ncol = 1)

```

![Correct Buffer and with Better with More Islands](buffer_island_example.pdf)
